# Complete protocols {#complete-protocols}

This chapter showcases a number of complete protocols for different kinds of data visualizations. Each protocol starts with the raw data and ends with a publication quality plot. The data is available from Github, just follow [this link](http://hmpg.net). In the same folder you can find the R markdown files for each protocol. To reproduce the data visualizations you can either take the data and follow the instructions in the chapter. Alternatively you can download the R markdown file (and the data) and run it step by step. The R markdown files can be used as a starting point to apply the same visualization to your own data.

``` {r child = 'Protocols/protocol_1.Rmd'}
```


``` {r child = 'Protocols/protocol_2.Rmd'}
```


``` {r child = 'Protocols/protocol_3.Rmd'}
```


``` {r child = 'Protocols/protocol_4.Rmd'}
```


``` {r child = 'Protocols/protocol_5.Rmd'}
```



#### In progress {-}

This chapter is 'in progress' and I plan to add more protocols in the future. A list of ideas:

* Heatmap-style representation of timeseries data
* Waterfall plot (timeseries)
* 96-well plate data
* Dose-reponse curves
* Plot data from a Google form
* Polar plot with Timestamp data
* Plotting grouped data 
* Annotated timeseries (optogenetics)
* Volcano plot
* Animated plot
* Multipanel figure (KTR data)
* correlation/regression/Bland-Altman
* Plots of Effect size (differences)
* Permutated p-values (not strictly dataViz)
* Bootstrapped confidence intervals


