---
title: "Protocol 3 - step-by-step"
subtitle: "With flipbookr and xaringan"
author: "Joachim Goedhart"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, hygge, ninjutsu]
    nature:
      ratio: 16:9
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r, include = F}
# This is the recommended set up for flipbooks
# you might think about setting cache to TRUE as you gain practice --- building flipbooks from scratch can be time consuming
knitr::opts_chunk$set(fig.width = 6, message = FALSE, warning = FALSE, comment = "", cache = F)
library(flipbookr)
library(tidyverse)
```



# Hi!

--

This is a 'flipbook' that show how to build a plot step-by-step!

--

Flipbooks are code-movies.  The frames of the movies are slides. 


<!-- get a flipbook version of the my_cars code chunk - pauses are where parentheses are balanced -->
---

`r chunk_reveal("protocol_3")`

```{r protocol_3, include = FALSE}
read.csv("protocol_3.csv") %>%
  ggplot(., aes(x=time, y=Roundness, group=Cell)) +
  geom_line() +
  facet_wrap(~Cell) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1.0)) + 
  scale_x_continuous(expand = c(0,0), limits = c(0, 300)) +
  theme_minimal(base_size = 14) + 
  theme(strip.background = element_blank(),
               strip.text = element_blank(),
               plot.caption = element_text(color = "grey80"),
               #Remove grid
               panel.grid.major = element_blank(),
               panel.grid.minor = element_blank()
        ) + theme(        
        #Remove axis labels
        axis.text = element_blank()
        )

```

---
class: inverse, center, middle


# Next up:  refer to same chunk, but use user defined breakpoints, and display output only






<!-- get another flipbook version of the my_cars code chunk, this time at the defined breaks and only showing output -->
---

`r chunk_reveal("my_cars", break_type = "user", display_type = "output", title = "### user defined break! using #BREAK message")`



---
class: inverse, center, middle

# The End




<!-- adjust font size in this css code chunk, currently 80 -->

```{css, eval = TRUE, echo = FALSE}
.remark-code{line-height: 1.5; font-size: 80%}

@media print {
  .has-continuation {
    display: block;
  }
}

code.r.hljs.remark-code{
  position: relative;
  overflow-x: hidden;
}


code.r.hljs.remark-code:hover{
  overflow-x:visible;
  width: 500px;
  border-style: solid;
}
```



---

# Notes: 

- This minimal flipbook assumes you are online when you build and view.  It is going to find the most recent remark.js info online.

- xaringan/remark.js slideshows are not self contained, single file slide shows.  Make sure you ship all the associated files that are created to display your flipbooks/slideshows.  I like to push .html and associated files to github repository and use Github Pages to host for beyond local sharing. :-)

