---
author: "Joachim Goedhart"
date: "`r Sys.Date()`"
output: html_document
---

## Protocol 23 - Plotting multiple conditions side-by-side {#protocol-23}

In this protocol, we plot data that is gathered on 'cellular brightness' as described in [this blog post](https://blog.addgene.org/a-practical-approach-to-choosing-the-brightest-fluorescent-protein). In this specific set, different proteins are measured before (pre) and after (post) saturating the cells with calcium. The dataset has brightness values per cell and there are quite some data-points that are gathered from two independent experiments. This data was first published [here](https://doi.org/10.1101/2024.11.05.622032
) and the code below can be used to replicate figure 4 of the preprint.

First the necessary packages are activated:

```{r}
library(tidyverse)
library(ggbeeswarm)
```

Next, the data is loaded as CSV into a dataframe:

```{r}
df <- read.csv("data/ratio_cells_selection_for_JG_2.csv")
head(df)
```

There are more data than I want to show, so I filter the data based on the name in the column `construct`, and this cleaned dataset is saved:

```{r}
df %>% filter(construct %in% c("EGFP", "G-Ca-FLITS", "mTq2_T203Y","GCaMP6s", "jGCaMP7c", "GCaMP3", "G-GECO1.1")) %>% write.csv("data/ratio_cells_FLITS_set.csv", row.names = F)
```

We can load the filtered data, which is a standard CSV file:

```{r}
df <- read.csv("data/ratio_cells_FLITS_set.csv")
head(df)
```

The column `glass` identifies the replicate, so we change the name to better reflect that:

```{r}
df <- df %>% rename(replicate=glass)
```

Let's also clean up some of the qualitative variables:

```{r}
df <- df %>% mutate(construct =
                      case_when(construct == "mTq2_T203Y" ~ "mTq2-T203Y",
                                TRUE ~ construct)
                    )

df <- df %>% mutate(across('replicate', str_replace, 'g', ''))

head(df)

```



The data is in tidy format. The quantitative variable that we want to plot is the `norm_ratio` per construct and state.
One way to achieve this by specifying different fill colors for the `state` and plotting the distributions with `geom_violin()`:

```{r}
ggplot(df, aes(x=construct, y=norm_ratio)) + geom_violin(aes(fill=state))
```

Another way is to do this with `facet_wrap()`:

```{r}
ggplot(df, aes(x=state, y=norm_ratio)) + geom_violin() + facet_wrap(~construct)
```

Here, I think the second option is much nicer, as it shows the different constructs as a label on top of the figure. Let's work on this plot to improve it. It makes more sense that the 'pre' condition is shown first:

```{r}
df <- df %>% mutate(state = fct_relevel(state, c("pre", "post")))
```

And, in the same way, I will group the true GFP-based probes (EGFP,GCaMP3,GCaMP6s,jGCaMP7c) and the Turquoise-based probes (G-Ca-FLITS,mTq2_T203Y):

```{r}
df <- df %>% mutate(construct = fct_relevel(construct, c("EGFP","G-GECO1.1", "GCaMP3","GCaMP6s","jGCaMP7c","G-Ca-FLITS","mTq2-T203Y")))
```

We can also show the individual datapoints according to their distribution by using `geom_quasirandom()` from the `{ggbeeswarm}` package:

```{r}
 ggplot(df, aes(x=state, y=norm_ratio, group = replicate)) + geom_quasirandom(alpha=0.3, size=2) + facet_wrap(~construct)
```

Let's compress the plot in the y-direction, showing all plots in a single row (and add some color):
We also 'compress' the plot, taking less space:

```{r}
p <- ggplot(df, aes(x=state, y=norm_ratio, group = replicate, fill=construct, color=construct)) + geom_quasirandom(alpha=0.3, size=2) + facet_wrap(~construct, nrow = 1)
p <- p + theme_bw(base_size = 16) + theme(aspect.ratio = 4)
p
```

To display the median value of the replicates we add a larger dot showing the median and add a line to connect these paired data:

```{r}
p <- p +
  stat_summary(fun = median, geom="line", color="black", linewidth=1) +
  stat_summary(fun = median, geom='point', size=6, shape=21, color="black", alpha=0.5)
p
```

Let's edit the plot to adjust the theme, labels and layout:

```{r}
p <- p + scale_color_manual(values = c('darkgreen','darkgreen','darkgreen','darkgreen','darkgreen','darkseagreen4','darkseagreen4'))
p <- p + scale_fill_manual(values = c('darkgreen','darkgreen','darkgreen','darkgreen','darkgreen','darkseagreen4','darkseagreen4'))


p <- p + guides(fill = "none",
                color = "none"
                )

p <-
  p + labs(
    title = "Cellular brightness of green Calcium biosensors",
    subtitle = "pre and post ionomycin addition",
    tag = "Protocol 23",
    y = "normalized brightness"
  ) +
  theme(panel.grid = element_blank()) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1.3))

p
```

To save the plot:

```{r }
png(file=paste0("Protocol_23.png"), width = 4000, height = 3000, units = "px", res = 400)
 p
dev.off()
```
