
---
author: "Joachim Goedhart"
date: "`r Sys.Date()`"
output: html_document
---

## Protocol 8 - Plotting data in a 96-wells layout {#protocol-8}

This protocol displays data from a 96-wells plate in the same format as the plate. This data visualization of 96-well plate data is also used in the [plotXpress app](https://doi.org/10.12688/f1000research.73641.1).

We start by loading a package that we need:

```{r }
library(tidyverse)
```

A good example of data wrangling related to 96-wells data is given in [protocol 4](#protocol-4). Here we start from the tidy data that was the result of data tiyding in the section [Data in 96-wells format](#Data-in-96-wells-format):

```{r }
df_8 <- read.csv("df_tidy_wells.csv")
head(df_8)
```

We can construct a basic plot where each well is represented by a large dot and the color used to fill the dot represents the intensity. This is much like a heatmap, but it uses a dot instead of a tile:

```{r}
ggplot(data=df_8, aes(x=column, y=row)) +
geom_point(aes(color=Intensity), size=10)
```
There is a couple of things that need to be fixed. First, we invert the row names, to put the row 'A' at the top. In addition, the column names are depicted as factors (this can also be achieved by mutating the dataframe):

```{r}
p <- ggplot(data=df_8, aes(x=as.factor(column), y=fct_rev(row))) +
  geom_point(aes(color=Intensity), size=10)
p
```

Let's change the color palette that displays intensity to viridis. The numbers in the legend are high and that is why they are difficult to read quickly. So we use the `scales` package to use the scientific notation:

```{r}
  p <- p + scale_x_discrete(position = "top") +
  scale_color_viridis_c(label = scales::scientific)
p
```

In my opinion, the scientific notation does not work well, so let's try to use commas as a thousands separator:

```{r}
  p <- p + scale_x_discrete(position = "top") +
  scale_color_viridis_c(label = scales::comma)
p
```

This is a bit better, but the numbers need to be right aligned:

```{r}
p <- p + theme(legend.text.align = 1)
p
```
This looks better, now add the labels (and we skip the labels for the x- and y-axis):

```{r }
p <-
  p + labs(
    title = "Readings from a 96-wells plate",
    subtitle = NULL,
    x = NULL,
    y = NULL,
    caption = "\n@joachimgoedhart\nbased on data from Brandorff et al., DOI: 10.1101/2021.07.08.451595",
    color= 'Intensity [a.u.]',
    tag = "Protocol 8"
  ) 
p
```

Set the theme and font size:

```{r }
p <- p + theme_light(base_size = 14)
```
Note that this overrides any previously defined modifications of the `theme()` function, such as the alignment of the legend labels. So we need to define this again.

```{r}
p <- p + theme(legend.text.align = 1)
```

We also adjust the theme settings for the other elements:

```{r }
p <- p + theme(plot.caption = element_text(color = "grey80"),
               plot.subtitle = element_text(color = "grey50", face = "italic"),
               #Remove the grid
               panel.grid.major = element_blank(),
               panel.grid.minor = element_blank()
)
```

Let's look at the result:

```{r }
p
```

To save the plot as a png file:

```{r }
png(file=paste0("Protocol_08.png"), width = 3000, height = 2000, units = "px", res = 400)
 p + coord_fixed()
dev.off()
```

