---
author: "Joachim Goedhart"
date: "`r Sys.Date()`"
output: html_document
---

## Protocol 1 - Spectra of fluorescent proteins {#protocol-1}

This protocol describes how you can turn a csv with spectral data that is obtained from FPbase.org into a plot of those spectra. First, we load the required package:

```{r }
library(tidyverse)
```

For this data visualization, I selected several spectra from fluorescent proteins at FPbase.org: [https://www.fpbase.org/spectra/?s=1746,6551,101,102,123,124,1604,1606&showY=0&showX=1&showGrid=0&areaFill=1&logScale=0&scaleEC=0&scaleQY=0&shareTooltip=1&palette=wavelength](https://www.fpbase.org/spectra/?s=1746,6551,101,102,123,124,1604,1606&showY=0&showX=1&showGrid=0&areaFill=1&logScale=0&scaleEC=0&scaleQY=0&shareTooltip=1&palette=wavelength)

The data was downloaded in CSV format (by clicking on the button in the lower right corner of the webpage) and renamed to 'FPbase_Spectra_4FPs.csv'.
We read the data from the CSV by using the `read_csv()` function. This function is part of the tidy verse and loads the data as a tibble. It also guesses type of data for each column. To hide that information, we use `show_col_types = FALSE` here.

```{r }
df_raw <- read_csv("data/FPbase_Spectra_4FPs.csv", show_col_types = FALSE)
```

Let's briefly look at what we have loaded:

```{r message=FALSE, warning=FALSE}
glimpse(df_raw)
```

The data needs conversion to a tidy format before plotting. Since we have a single continuous data column with Wavelength information that is used for the x-axis, it is excluded from the operation: 

```{r }
df_1 <-
  pivot_longer(
    df_raw,
    cols = -Wavelength,
    names_to = "Sample",
    values_to = "Intensity"
  )
```

There are several rows that have NA values for Intensity and this is how to get rid of that:

```{r }
df_1 <- df_1 %>% drop_na(Intensity)
```

The column 'Sample' has labels for the fluorescent protein and the type of spectrum. We can separate that column into two different columns that we name 'Fluorescent Protein' and 'Spectrum':

```{r }
df_1 <- df_1 %>%
  separate(Sample, c("Fluorescent Protein", "Spectrum"), sep = " ")
```

Let's do a first attempt and plot the data:

```{r }
ggplot(
  data = df_1,
  aes(x = Wavelength, y = Intensity, color = `Fluorescent Protein`)
) +
  geom_line(aes(linetype = Spectrum), size = 1)
```

This looks pretty good already. Now let's change the order of the fluorescent proteins to their order in the plot:

```{r }
df_1 <- df_1 %>%
  mutate(`Fluorescent Protein` = forcats::fct_relevel(
    `Fluorescent Protein`,
    c("mTurquoise2", "mNeonGreen", "mScarlet-I", "miRFP670")
  ))
```

The data is in the right shape now, so let's save it:

```{r}
df_1 %>% write.csv("protocol_1.csv", row.names=FALSE)
```

We define the plot object and add an extra geometry, `geom_area()` to fill the area under the curves:

```{r }
p <-
  ggplot(
    data = df_1,
    aes(
      x = Wavelength, y = Intensity,
      fill = `Fluorescent Protein`
    )
  ) +
  geom_line(aes(linetype = Spectrum),
    size = 0.5, alpha = 0.5
  ) +
  geom_area(
    aes(linetype = Spectrum),
    color = NA,
    position = "identity",
    size = 1,
    alpha = 0.5
  )
```

Let's check the result:

```{r}
p
```

Next, we set the limits of the axis and force the y-axis to start at 0

```{r }
p <-
  p + scale_y_continuous(expand = c(0, 0), limits = c(0, 1.1)) + 
  scale_x_continuous(expand = c(0,0), limits = c(350, 810))
```

Add labels:

```{r }
p <-
  p + labs(
    title = "Spectra of Fluorescent Proteins",
    x = "Wavelength [nm]",
    y = "Normalized Intensity [a.u.]",
    caption = "@joachimgoedhart\n(based on data from FPbase.org)",
    tag = "Protocol 1"
  )
```

Modify the layout by adjusting the theme. Comments are used to explain effect of the individual lines of code:

```{r }
p <-
  #Set text size
  p + theme_light(base_size = 14) + theme(
    plot.caption = element_text(
      color = "grey80",
      hjust = 1
    ),
    #Remove grid
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    #Set position of legend
    legend.position = "top",
    legend.justification = "left"
    #Define the legend layout
  ) + guides(
    linetype = "none",
    fill = guide_legend(title = NULL, label.position = "right")
  )
```

```{r}
p
```

We are almost there, except that the colors of the plot do not match with the natural colors of the fluorescent proteins. Let's fix that by defining a custom color palette. The order of the colors matches with the order of the fluorescent proteins that was defined earlier:


```{r }
custom_colors <- c("blue", "green", "orange", "red")
```

To apply the custom colors to the filled area:

```{r }
p <- p + scale_fill_manual(values = custom_colors)
```

This is the result:

```{r }
p
```

To save this plot as a PNG file:

```{r }
png(file=paste0("Protocol_01.png"), width = 4000, height = 3000, units = "px", res = 400)
  p
dev.off()
```
