---
author: "Joachim Goedhart"
date: "`r Sys.Date()`"
output: html_document
---

## Protocol 27 - Raincloud plot {#protocol-27}

```{r}
library(tidyverse)
```

The donut chart that we've seen in protocol 26 summarizes a rich dataset in a single value. While this simplification can be a powerful way to communicate results, we may be interested in the underlying data. A comprehensive way to show all data and its distribution is the raincloud plot. This is a recent innovation that was reported by  [Micha Allen and colleagues](https://doi.org/10.12688/wellcomeopenres.15191.2). We will use the same dataset as for protocol 26 to generate a raincloud plot and this will nicely contrast with the donut chart. Let's again load the data on the percentage of HeLa cells in the S-phase:

```{r}
df_S <- read.csv("data/results_S-phase-5yrs.csv") %>% filter(Group != "D")
head(df_S)
```

This data is cleaned and it has information on the year (ranging from 2021-2025), the group (A/B/C/D), whether the data was analysed by hand or automatically and the percentage of S-phase cells that was detected. There is quite some data on the manual versus automated analysis and I want to see the distribution of the data and evaluate whether these approaches give similar results.

Let's first look at the distributions:

```{r}
ggplot(df_S, aes(x=S_phase, fill=Analysis)) +
  geom_density() +
  facet_wrap(~Analysis)
```

We do not see the actual data, and we can add these with `geom_rug()`:
```{r}
ggplot(df_S, aes(x=S_phase, fill=Analysis)) +
  geom_density() + geom_rug() +
  facet_wrap(~Analysis)
```

This gives some idea of the data, but in this plot, the emphasis is on the distribution. To better visualize the distribution and the data, one may combine a jittered dotplot with a violinplot. Note that we generate a standard jitter+violin plot, and then we rotate it to make it comparable to the distributions in the plot above:

```{r}
ggplot(df_S, aes(x=Analysis, y=S_phase, fill=Analysis)) +
  geom_violin() + geom_jitter() + coord_flip()
```


```{r}

```

The violinplot shows the distribution, but it does so twice. To show it once, a new `geom` has been defined, i.e. `geom_flat_violin()`. This is not (yet) supported by the `{ggplot2}` package and therefore we need to load it separately:

```{r}
source("https://raw.githubusercontent.com/JoachimGoedhart/SuperPlotsOfData/refs/heads/master/geom_flat_violin.R")
```


```{r}
ggplot(df_S, aes(x=Analysis, y=S_phase, fill=Analysis)) +
  geom_flat_violin() + geom_jitter() +coord_flip()
```

Now, the distributions look like a cloud, but we need to reduce the jitter, and to move the clouds up:

```{r}
ggplot(df_S, aes(x=Analysis, y=S_phase, fill=Analysis, color=Analysis)) +
  geom_flat_violin(position = position_nudge(x = .15, y = 0), width=0.8) +
  geom_jitter(width = 0.1) + coord_flip()
```


The original raincloudplots include a boxplot, but that's a bit too cluttered in my opinion:

```{r}
ggplot(df_S, aes(x=Analysis, y=S_phase, fill=Analysis, color=Analysis)) +
  geom_flat_violin(position = position_nudge(x = .15, y = 0), width=0.8) +
  geom_boxplot(position = position_nudge(x = 0.15, y = 0), width = 0.05, color="black", outlier.shape = NA, fill=NA) +
  geom_jitter(width = 0.1) + coord_flip()
```

So let's define a raincloudplot without the box and do some styling. We will also change the order of the the methods:

```{r}
p <- ggplot(df_S, aes(x=fct_rev(Analysis), y=S_phase, fill=Analysis, color=Analysis)) +
  geom_flat_violin(position = position_nudge(x = .15, y = 0), width=0.8) +
  geom_jitter(width = 0.1) + coord_flip() +
  theme_light(base_size = 16)
p
```

Deafault colors are OK-ish, but let's use the same colors we've used for the donut chart in protocol 26:
```{r}
p <- p + scale_fill_manual(values=c("darkorange2", "dodgerblue3")) +
  scale_color_manual(values=c("darkorange2", "dodgerblue3"))

p
```



```{r}
p <- p + theme(legend.position = "none") +
    theme(plot.caption = element_text(color = "grey80", hjust = 1)) +
    labs(title = "The percentage of cells in S-phase...",
    subtitle = "...determined by two methods",
    x="Analysis",
    y="S-phase [%]",
    caption = "@joachimgoedhart | data submitted by students",
    tag = "Protocol 27"
    )

p
```

Saving the plot:

```{r }
png(file=paste0("Protocol_27.png"), width = 4000, height = 3000, units = "px", res = 400)
 p
dev.off()
```

```{r}
detach(package:plyr) #needed to get rid of this package which was required for the flat violin
```

