
---
author: "Joachim Goedhart"
date: "`r Sys.Date()`"
output: html_document
---

## Protocol 6 - Heatmap style visualization of timelapse data  {#protocol-6}

Lineplots are typically used to plot data from timeseries (also known as longitudinal data). However, in case of many samples/objects, this may result in a cluttered data visualization. To solve this, the same data can be presented as a heatmap, where every row is an object and the response is coded as a color. A downside is that it is less quantitative as it is difficult to 'read' the numbers from this kind of visualization. Still, it is a powerful visualization to plot a lot of data and show its dynamics and heterogeneity. The data and visualization is originally published by [Chavez-Abiega et al. (2021)](https://www.biorxiv.org/content/10.1101/2021.07.27.453948v3)

The multipurpose `{tidyverse}` package is used for data wrangling and plotting:

```{r }
library(tidyverse)
```

A CSV file with the data is loaded. Since the file is a couple of Megabytes, we use the faster `fread()` function from the package `data.table`:

```{r }
library(data.table)
df_S1P <- data.table::fread("data/ERK_repsonse_to_S1P.csv") 
head(df_S1P)
```

The column 'CN_ERK' has the data on the activity that we will plot over time. Each number in the 'Unique_Object' column reflects an individual cell measurement and so we can use that to group the measurements using `group_by(Unique_Object)`. We subtract the average baseline activity from each trace by subtracting the data acquired at the first 5 timepoints: `CN_ERK[1:5]`. The data is stored in a new column with normalized ERK activity data 'ERKn':

```{r }
df_sub <- df_S1P %>% group_by(Unique_Object) %>% arrange(Time_in_min) %>% mutate(ERKn=CN_ERK-mean(CN_ERK[1:5])) %>% ungroup()
```

Around Timepoint 23 (minutes), the ligand was added. To set this time point to zero, we subtract a value of 23 (minutes) from each point:

```{r }
df_sub$Time_in_min <- df_sub$Time_in_min-23
```

The column 'Unique_Object' that identifies the individual cells contains natural numbers, but these need to be treated as qualitative data. Therefore, we change the data type in this column to a factor with `as.factor()`:

```{r }
df_sub <- df_sub %>% mutate(Unique_Object=as.factor(Unique_Object))
```

To order objects, we need to order 'Unique_Object' according to something. That something can be the maximum value of the Erk activity:

```{r }
df_sub <- df_sub %>% mutate(Unique_Object = fct_reorder(Unique_Object, ERKn, max))
```

Plot the data in heatmap style. We use theme_void here to focus only on the data and we will deal with styling of the axes and labels later:

```{r }
ggplot(df_sub, aes(x=Time_in_min, y=Unique_Object,fill=ERKn)) +
  geom_tile() + theme_void()+
  scale_fill_viridis_c() 
```

Can we also sort the data based on something else? Definitely, but it requires a bit of understanding of functions. Previously, we used the maximum value. This is defined by the function `max`, which takes the maxium value from a vector of numbers. Let's look at an artificial example:

```{r }
x <- c(1,3,4,5,6,7,9)
max(x)
```

 Other functions that take a vector as input and return a single value as output can be used. Other existing examples are `mean()`, `sum()` and `min()`. We can also define a function:

```{r }
second <- function(x) {x[2]}
second(x)

one_but_last <- function(x) {x[length(x)-1]}
one_but_last(x)
```

We can use the new function to sort the dataframe: 

```{r }
df_sub <- df_sub %>% mutate(Unique_Object = fct_reorder(Unique_Object, ERKn, one_but_last)) 

ggplot(df_sub, aes(x=Time_in_min, y=Unique_Object,fill=ERKn)) +
  geom_tile() + theme_void()+
  scale_fill_viridis_c() 
```

If we want to sort on the sum of the top five values we can define a function:

```{r }
top_five <- function(x) {sum(tail(sort(x),5))}
```

But we can also directly implement the function in the `fct_reorder()` function:

```{r }
df_6 <- df_sub %>% mutate(Unique_Object = fct_reorder(Unique_Object, ERKn, function(x) {sum(tail(sort(x),5))}))


```

Let's save this data:
```{r}
df_6 %>% write.csv("protocol_6.csv", row.names=FALSE)
```


```{r}
p <-  ggplot(df_6, aes(x=Time_in_min, y=Unique_Object,fill=ERKn)) + geom_tile() +
  # scale_x_continuous(breaks=seq(0,60, by=15), labels=seq(0,60, by=15), limits=c(-8,60))  +
  scale_fill_viridis_c(na.value="black", limits = range(-0.1,1.5))
```

Let's look at the plot

```{r }
p
```

Add labels:

```{r }
p <-
  p + labs(
    title = "ERK activity is increased by S1P",
    subtitle = "and varies strongly between cells",
    x = "Time after addition of S1P (min)",
    y = "Cells sorted according to response",
    caption = "@joachimgoedhart | data from Chavez-Abiega, DOI: 10.1101/2021.07.27.453948",
    tag = "Protocol 6",
    fill= "ERK C/N ratio"
  )
```

The `theme_void()` would be close to what we want as a theme, but I prefer to start from `theme_light` and remove the redundant features (grids and y-axis labels):
  
```{r}  
p <- p + theme_light(base_size = 14) +
  
  theme(plot.caption = element_text(color = "grey80", hjust = 2.0),
        plot.title = element_text(hjust = 0.1, margin = margin(t=10)),
        plot.subtitle = element_text(hjust = 0.1, margin = margin(t=2, b=5)),
        # Remove background
        panel.background = element_blank(),
        # Remove borders
        panel.border = element_blank(),
        # Remove grid
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        # Remove text of the y-axis
        axis.text.y = element_blank(),
        # Remove ticks on y-axis
        axis.ticks.y = element_blank(),
        # Remove label of y-axis
        axis.title.y = element_blank(),
        # Make x-axis ticks more pronounced
        axis.ticks = element_line(colour = "black")
  )

p
```

The caption does not look good in this plot, but it has been optimized to look good in the saved PNG. To get a proper aligned caption in the Rmd you may need to optimize the hjust value in `theme((plot.caption = element_text()))`

To save the plot as PNG:

```{r}
 png(file=paste0("Protocol_06.png"), width = 4000, height = 3000, units = "px", res = 400)
   p
 dev.off()
```

