---
author: "Joachim Goedhart"
date: "`r Sys.Date()`"
output: html_document
---

## Protocol 3 - small multiples of time courses {#protocol-3}

This protocol displays a number of different timecourses as 'small multiples'. Small multiples, as the name suggests, displays many small plot separately as a stamp collection. By stressing the data, rather than the labels and grids, this can be a powerful visualization strategy.

The data is taken from a publication by [Arts et al. (2021)](https://doi.org/10.3389/fimmu.2021.667213) and we recreate figure panel 1F. The original figure is in small multiple format, but we tweak it a bit more to increase the focus on the data.

Let's first load the necessary package:
```{r }
library(tidyverse)
```

The data comes from an excel file:

```{r }
df_raw <- readxl::read_excel("Data_Arts_Circularity.xlsx")
head(df_raw)
```

It is in a wide format, so we need to make it tidy. The parameter that was measured over time is the 'roundness' of cells:

```{r }
df_3 <-
  pivot_longer(
    df_raw,
    cols = -time,
    names_to = "Cell",
    values_to = "Roundness"
  )
```

The data is in the right shape now, so let's save it:

```{r}
df_3 %>% write.csv("protocol_3.csv", row.names = FALSE)
```


First we create a line plot of all the data:

```{r}
p <- ggplot(df_3, aes(x=time, y=Roundness, group=Cell)) + geom_line()
p
```

With the facet_wrap() function, we turn this into a small multiple:

```{r}
p <- p + facet_wrap(~Cell)
p
```


Set the limits of the axis and force the y-axis to start at 0

```{r }
p <-
  p + scale_y_continuous(expand = c(0, 0), limits = c(0, 1.0)) + 
  scale_x_continuous(expand = c(0,0), limits = c(0, 300))

```

Use a minimal theme and remove the strips and grid to increase focus on the data:

```{r}
p <- p + theme_minimal(base_size = 14)

p <- p + theme(strip.background = element_blank(),
               strip.text = element_blank(),
               plot.caption = element_text(color = "grey80"),
               #Remove grid
               panel.grid.major = element_blank(),
               panel.grid.minor = element_blank()
)
p
```

I do not like the repeated axis for the different plots. We can remove those:
```{r}
p <- p + theme(        
        #Remove axis labels
        axis.text = element_blank()
        )
p
```

This is a very minimal plot, focusing entirely on the data. It may work well, but it is informative to add some information about the scaling of the x- and y-axis. To achieve this, I add lines to the lower left plot, which correspond to the data of 'neutro 6' (you can see this in the small multiple plot where each plot was labeled). I define a new dataframe with the x- and y-scale for 'neutro 6' to do just that:

```{r }
ann_line<-data.frame(xmin=0,xmax=300,ymin=0,ymax=1,
                     Cell=factor("neutro 6",levels=c("neutro 6")))
ann_line
```

This dataframe can now be used to draw two lines with `geom_segment()`:

```{r }

p <- p +
  #Vertical line
  geom_segment(data=ann_line, aes(x=xmin,xend=xmin,y=ymin,yend=ymax), size=2, color='grey40') +
  #Horizontal line
  geom_segment(data=ann_line, aes(x=xmin,xend=xmax,y=ymin,yend=ymin), size=2, color='grey40') +
  NULL

p

```

The plot is now in black and white which gives it a strong contrast. We can make it a bit more soft and pleasant to look at by changing to shades of grey. Also, the labels of the axes are moved next to the lines:

```{r}
p <- p +
  theme(panel.background = element_rect(fill='grey98', color=NA),
        panel.border = element_rect(color='grey90', fill=NA),
        axis.title.x = element_text(size=14, hjust = 0, color='grey40'),
        axis.title.y = element_text(size=14, vjust = 0, hjust=0, angle = 90, color='grey40'),
  ) 
```

Finally, we add a title, caption, and labels (and a scale in brackets):

```{r }
p <-
  p + labs(
    title = "Changes in the shape of migrating cells over time",
    x = "Time [300s]",
    y = "Circularity [0-1]",
    caption = "@joachimgoedhart\n(based on data from Arts et al., DOI: 10.3389/fimmu.2021.667213)",
    tag = "Protocol 3"
  ) 
```



```{r}
p
```


```{r }
png(file=paste0("Protocol_3.png"), width = 4000, height = 3000, units = "px", res = 400)
 p
dev.off()
```

