---
title: "Measurements from Google form"
author: "Joachim Goedhart"
date: "12/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Plotting data that was harvestd with a Google form

The goal of this script is to read data that was generated by a Google form and transform it into a tidy format. We need the tidyverse package:

```{r }
library(tidyverse)
```

The data from the Google form are stored in a sheet that was published on the web as a CSV. This was achieved by 'File > Share > Publish to web' and selecting to 'Publish' as 'Comma-separated values'.
After clicking on 'Publish' a link is generated that can be read as a csv. Any empty cells are converted to `NA`:

```{r}
df_sheet <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSc-nI1-s_u-XkNXEn_u2l6wkBafxJMHQ_Cd3kStrnToh7kawqjQU3y2l_1riLigKRkIqlNOqPrgkdW/pub?output=csv", na.strings = "")
head(df_sheet)
```

There are a number of issues with this data. The first 5 rows are empty. I deleted the values in these rows, since they were dummy values from testing the form. Let's remove rows without data:

```{r}
df_sheet <- df_sheet %>% na.omit()
```

The header names are still in the original language (Dutch) and the third and fourth header names with the relevant data are very long. Let's change the names:

```{r}
colnames(df_sheet) <- c("Timestamp", "Group", "Cell", "Nucleus")
```

All the length measurements are distributed over two columns, let's make it tidy and move it to one column:

```{r }
df_tidy <-
  pivot_longer(
    df_sheet,
    cols = -c("Timestamp", "Group"),
    names_to = "Sample",
    values_to = "Size"
  )
```

The column with measurements contains up to 10 length measurements. The values are separated by a comma and often a space. First, the space is removed and then each row is split into multiple rows, where each row has a single measurement. The comma is used to separate the values:

```{r}
df_tidy <- df_tidy %>% mutate(Size = gsub(" ", "", Size)) %>% separate_rows(Size, sep=",") 
```

The 'Size' column values are converted to the class numeric. The advantage is that anything that is not a number will be converted to `NA`. These non-numeric data can stem from incorrectly enetered data. Examples of these incorrect data are numbers with units or with wrong separators.

```{r}
df_tidy <- df_tidy %>% mutate(Size = as.numeric(Size))
```

Finally, we can filter any values that fall outside the range of expected values (and this will also remove the cells with `NA`):

```{r}
df_size <- df_tidy %>% filter(Size>0 & Size<1000)
head(df_size)
```


The dataframe is cleaned and tidy and so it is ready for visualization. The primary information is the distribution of size measurements for the measurements of cells and cell nuclei. This can be achieved by using `geom_density()`. Given the range of sizes, it makes sense to use a log10 scale for the x-axis:

```{r}
ggplot(df_size, aes(x=Size, fill=Sample))+geom_density(alpha=.8) +
  scale_x_log10()
```

The `geom_density()` produces a smooth distribution of the data. I actually prefer to see the real data and that's why we turn to `geom_histogram`. I also like to split the data according to the sample and groups and this can be achieved with `facet_grid()`. Here, we use the `labeller()` function to combine the column name 'Group' with the the factor it presents, as this makes it easier to understand what is plotted:

```{r}
ggplot(df_size, aes(x=Size, fill=Sample))+geom_histogram(alpha=.8) +
  facet_grid(Sample~Group, labeller = labeller(Group=label_both)) +
  scale_x_log10()
```

The use of color is redundant, since the data is split in Cell and Nucleus anyway, so we remove that. The difference between groups is not large and here I chose to combine the data for all the groups: 

```{r}
p <- ggplot(df_size, aes(x=Size)) +
  geom_histogram(alpha=.5) +
  facet_grid(~Sample) +
  scale_x_log10()
p
```

The data is presented in a nice way, now we can improve the axis labels and add a title and caption:

```{r }
p <-
  p + labs(
    title = "Distribution of size measurements",
    subtitle = "measurements on human cells and nuclei of cells",
    x = "Size [µm] - log scale",
    y = "Count",
    caption = "(data submitted by 4 groups of students)",
    tag = "Protocol 10"
  )
```

Optimizing the layout:

```{r }
p <-
  #Set text size
  p + theme_light(base_size = 16) +
  theme(
    plot.caption = element_text(
      color = "grey80",
      hjust = 1
    ),
    axis.text.x = element_text(
      margin = margin(t=8)
    )
  )
```

```{r}
p + 
  #Force the y-axis to start at zero
  scale_y_continuous(expand = c(0, NA), limits = c(0,180)) +
  #Apply a logarithmic scale to the x-axis
  scale_x_log10(breaks = c(1,10,100)) +
  #Remove minor gridlines
  theme(panel.grid.minor = element_blank()) +
  #Add ticks to the bottom, outside
  annotation_logticks(sides="b", outside = TRUE) +
  #Needed to see the tcks outside the plot panel
  coord_cartesian(clip = "off")
```


To save the plot as a PNG file:
```{r }
png(file=paste0("Protocol_10.png"), width = 3000, height = 2000, units = "px", res = 400)
  p
dev.off()
```


